Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

### 2020-11-01 doc_arXiv/response_to_reviewers.Rmd

---
date: "`r Sys.Date()`"
title: "Response to reviewers"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
---

# Reviewer 1

__Comment__ This article assessed the impact of missing data imputation strategies and how they impacted accuracy of risk predictions. This was conducted on a case study dataset assessing mortality and transplant from a cardiac registry dataset, INTERMACS. It explains the motivation for the study in a nice two pronged approach: (1) the importance of the INTERMACS registry, how it is used, and the potential for better use of the available data to save lives, (2) the problematic approach of using mean imputation. The paper provides a thorough method to evaluate imputations, which I think could be used as a framework for future researchers. It is encouraging to see this kind of paper in this journal, as this kind of research, evaluating missing data strategies (in this case, imputation), are in my experience not frequently seen by this audience. It is also noteworthy that the authors have gone to efforts to ensure reproducibility of the work by providing the code and data in the repository (although the link was not made public yet, as far as I can tell). The paper has fantastic potential for impact, and will hopefully help change the way that other researchers and practitioners handle missing data to administer life saving treatment. Overall, the article is excellent

__Response__ We thank the reviewer for their kind words and helpful feedback. We apologize for initially keeping our Github repository private, this was an oversight on our part. The Github repository with all of our analysis code is now publicly available at https://github.com/bcjaeger/INTERMACS-missing-data.

<br>

__Comment__ While I have a few suggestions, these are quite minor. My main aim in this feedback is to help improve some aspects of communication from the authors to the reader.

__Response__ We appreciate the helpful suggestions that the reviewer has provided and provide a point-by-point response below.

## Handling vs imputing missing values

The paper describes exploring the 'handling' missing values, but from what I can tell, it is mostly about imputation strategies, and not handling. I would consider missing data handling methods to include decisions to remove observations, as well as 'logical' imputations (such as where missing values are identified as correctly being 0, or sections needing to be filled down).

I think that the authors could address this by referring to most cases of 'handle/ing' as 'imputation, and by also discussing missing data handling strategies such as listwise deletion. For example, in the 'Methods and Results' section of the Abstract, changing, 'depends on the strategy that was applied to handle missing data' to 'depends on the imputation strategy applied'. That said, it might be possible for the authors to cover deletion methods in their results, discussed in the next section (adding listwise deletion)

## Adding listwise deletion

On page 3 on the paragraph starting at line 4, the authors describe missing data handling. The typical behaviour is imputing to the mean - which is very bad, and it is great to see the authors addressing this. Do the authors have a comment on listwise deletion? It is almost always the wrong decision to perform, but I think it is worthwhile addressing in the paper, as it is unfortunately a practice that I think people might not be aware of as being very bad.

It could also serve as a baseline/bottom point for the simulation, to show the impact of not imputing. I would be interested in the authors thoughts on this - I don't think that adding listwise deletion as an additional part of the simulation is necessary for publication, but I do think it might improve the paper as it can communicate the dangers of deletion methods.

If the authors do not want to compare deletion methods to imputation I think it would be worthwhile to mention reasoning for this in the methods, or discussion section.

However, I do think that the term, 'missing data handling' encompasses the process of the data tidying and cleaning decisions that take place when deciding what to do with missing data, which happens to include things like deletion and imputation. If the authors do not wish to discuss deletion methods, then I think they should rephrase the mentions of 'missing data handling' to intead cover only 'imputation of missing data'. If this is the case, I think the authors should provide a sentence or two justifying why they chose not to explore deletion methods in addition to imputation.

## On Acronyms

There are quite a few acronyms in this paper. Would it be possible for the authors to reduce the number of acronyms? Sentences like 'Similar to the PH model, GBDT ensembles assume PH but apply a sequential modeling algorithm to develop RPEs' are a bit dense to parse. I wonder if it might be possible to remove some of the acronyms that are sparsely used.

I am also not entirely clear what an RPE is, exactly? This might be a term that is common in this area, so my apologies if this is perhaps misplaced, but how is it different to just 'prediction', or 'predicted values'? If it is an equation that produces a prediction, I'm not sure how this is different to a predicted value / model?

## Thoughts on Methods

The methods of this paper provide concise descriptions of the methodologies. One suggestion is to add a reference to Introduction to Statistical Learning for Gradient Boosted Decision Trees and a few of the other algorithms.

### Imputation methods

Lines 19-20 on page 6 states: 
'KNN imputation identifies k 'similar' observations for each case with at least one missing value'. 
Can the authors clarify here the distinction between an observation and a case?

### Missing patterns

I'm not sure why this section is here and how it relates to the previous sections. I suggest perhaps providing a sentence or two describing why this is important to cover. I understand it is mentioned later in the paper, but it seems this section is missing some level of context, and it could be improved by explaining how this relates to the paper, and why this matters for this research objective.

### 2.6 - computational details

SAS, R, and Python should be cited as well as relevant R and Python packages/modules used to produce results (such as ggplot2, which I am quite sure was used to create the graphics).

### A note on Tables and Figures

The caption text for the tables and figures could provide more detail to the reader. For example, Table 3 and Table 4 provide important information on the performance of a variety of tests, however it would be helpful to provide some guidance. One template I have found useful is to provide three parts to each caption. The first part is an overview, the second describing the contents, and the third being what we learn/take home message. Of course, this does not always hold true, for example, Table 2 (the typical 'table one' of the paper) that provides overview of the population - there is a lot of information here that is challenging to summarise, but if there are particular things to note about the population, they could be stated here.

I don't think captions need to be a whole paragraph, but something that I think will enhance the great work the authors have done already in the paper, so that the audience can absorb it and put it into practice. For example, tables 5 and 6 could be explained in more detail - what do we learn from this, which method is best and why?

Another example example, figure 2, could be expanded to state: 'Three Model evaluation metrics (mean with 95%CI errors), improve as the number of multiple imputed datasets increases. ...'. Figure 2 could be explained in more detail in the paper as well.

## Discussion

Page 20 line 2: I wonder if some of these references could be mentioned earlier in the paper in the introduction?

The paper demonstrates how good methods such as KNN are for imputation, however I think it might be just as critical to emphasize the importance of NOT using mean/median imputation methods.

Page 20 line 10 - 12: Can you tell me more about what you learn from the figure here?

Page 22 line 1: It is fantastic to see that the authors are sharing code. Obviously the data for the paper is likely very sensitive, but I wonder if simulated data is provided? I could not find the github repository they mention, perhaps it is currently set to private? Also the repository with the code should be deposited on zenodo (or Dryad, if Circulation is able to link the repository with Dryad) so a DOI can be provided and it can be cited.

One question I would be interested in seeing answered is what other types of data these approaches are suitable or unsuitable for in the field? So, if I am a medical practitioner with a database of heart attacks that have missing data, am I likely to get similar improvements? Should I use KNN imputation as well? My thoughts are yes, but perhaps there are key structures of the data that are relevant here.

The paper has great potential for impact - I can imagine a lot of registries using these approaches to impute missing data and getting improved treatment results, and I think that a key part of this is that the authors have made efforts to share their code. This could be emphasized more in the paper, so the reader knows that they too can use similar approaches with less effort.

Overall, a great paper.

Reviewer #2:

Review for CIRCCQO/2020/00707 - Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

In this study, the authors applied six strategies of handling missing values to develop three risk prediction equations (RPE), and compared the missing handling techniques on model performance and prediction accuracy. Overall, the manuscript is well-written. To increase the applicability of these missing handling techniques, I have summarized some areas that could be improved.

1. The authors used the missing handling techniques in their RPE development. These RPEs however are not yet established; and the performance of these RPE are not satisfied, for example, the c-index for mortality event is only around 65%. To increase the application of these missing handling techniques and not to distract the readers with the new RPEs, it would be better if the authors could compare these missing handling techniques using an existing prediction model to see if their approaches could improve the predictions.

2. Imputation accuracy: Can the authors compare the imputation accuracy of their techniques through either artificially inducing missing values to known data or simulating data or literature search? It's not critical to have imputation accuracy for prediction purpose (as authors indicated in the discussion), however, the imputation accuracy impacts model inference and will limit the use of certain missing handling techniques.

3. Page 9. Line 12-14 suggests up to 50 predictor variables were used, while page 13 Line 18 suggests 320 predictors with the one missing values exists in the registry. Could the authors clarify how they choose the candidate predictors for their RPE, as well as summarize the proportion of missingness and possible missing patterns for these predictors? For example, it is possible that the variables are missing from an INTERMACS center who failed to submit data. And the variables can also be missing for a certain time period when a center hasn't started their data participation. Under these "real world data" scenarios, can the author make recommendation in their missing handling techniques?

4. Page 12. Line 1-3. Please clarify the interpretation of scaled Brier score. Usually, the lower the Brier score is, the better the predictions are. It seems here, the scaled brier score is shown as the percentage relative to naïve RPE, the closer to 100% the better. In Table 3. Although the KNN (MI) methods improved the scaled Brier score, however, the scaled Brier scores are very closed to 0, which indicate low accuracy of a RPE.

5. Table 2. please clarify when modeling mortality events, how patients with transplants events and cessation were used, are they treated as censoring or are they dropped? Please explain the potential bias if the transplant patients were not included into the analysis for mortality events. The same question for modeling transplant events, how dead and cessations were handled in the cox model for transplant risk?

6. Table 3. The concordance is low, which indicates low model performance. The table title has a typo for "mortality".

7. Page 15. Line 3. Based on Table 3, for GBDT, the improvement in scale Brier score from KNN (MI) is 0.7 compared to the reference strategy. For CIF, the improvement from KNN (MI) is 0.8 when compared to the reference. These are slightly off in the text, which could be rounding issues. Please clarify.

8. Page 15 Line 6-8. From Table 3, for the CIF, stepwise PH and GBDT, the missing data strategies that provided the lowest calibration error were missForest, KNN (MI), and BR respectively. Please correct the order of the statement.

### 2020-11-01 doc_arXiv/doc_arXiv.Rmd
all tables and figs added

gt_latex <- function(gt_object, caption = 'a table', label = NULL){

  header = paste0(
    '\\begin{table} \n \\caption{',caption,'} \n'
  )

  if(!is.null(label)){
    header = paste0(header, '\\label{', label, '} \n')
  }

  gt_object %>% 
    as_latex() %>% 
    as.character() %>% 
    gsub("*", "",., fixed=T) %>% 
    gsub("\\\\ \n\\small \\\\ \n", "", ., fixed=T) %>% 
    gsub("\n\\large", "", ., fixed=T) %>% 
    gsub("\\captionsetup[table]{labelformat=empty,skip=1pt}", "", ., fixed=T) %>% 
    gsub("longtable", "table", ., fixed=T) %>% 
    gsub('\\begin{table}','\\begin{table} \n\\begin{tabular}', ., fixed=T) %>% 
    gsub('\\end{table}','\\end{tabular} \n \\end{table}', ., fixed=T) %>% 
    gsub('\\begin{table}', header, ., fixed=T) %>% 
    cat()

@@

drake::loadd(im, 
             times, 
             md_method_labels,
             md_type_labels,
             model_labels,
             outcome_labels,
             additional_missing_labels,
             tbl_characteristics,
             tbl_impute_accuracy)

@@

The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nation wide level. Currently, INTERMACS comprises over 20,000 patients who have received a MCS device. As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step for developing risk prediction models is handling missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial. While some missing values are related to data entry error, others may relate to instability of a patient’s circulatory status at the time of implant.  For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed. Patients without invasive cathaderizations  may be too ill (unstable) to tolerate the procedure. Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.

@@

Here we describe the INTERMACS Registry (Section \ref{subsec:intermacs}) and give a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning (Section \ref{subsec:inference_and_learning}). We then describe thirteen strategies to handle missing values (Section \ref{subsec:imputation}) and three strategies to develop risk prediction models after missing data have been imputed (Section \ref{subsec:modeling}). Next, we describe how risk prediction models and by extension the imputation strategy applied before training the model were evaluated in Section \ref{subsec:evaluation}. We outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}. Last, details on computation and software packages used for the current analysis are described in Section \ref{subsec:computing}

@@

INTERMACS is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS became an official Society of Thoracic Surgeons database.

@@

<!-- Though the outcome variable should be used when imputing predictor values for statistical inference, using outcome variables for this purpose in supervised learning projects can lead to issues in clinical implementation. For instance, suppose we seek to predict an outcome $Y$ for a patient in the clinical setting. If the patient is missing information for a predictor $X$ and our only method to impute missing values of $X$ leverages the observed value of $Y$, how should we impute $X$? If we do not already know $Y$, which would seem likely for clinical application of a prediction model, then we cannot impute $X$. On the other hand, if we already know $Y$, then we obviously do not need to predict it. Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice \cite{thomas2014pre}, we did not leverage outcome variables (\ie time until transplant, death, or last contact) to impute missing values of predictors in the current analysis. -->

@@

\paragraph{K-Nearest-Neighbors (KNN)} KNN imputation identifies $k$ `similar' observations (\ie neighbors) for each observation with a missing value in a given variable \cite{chen2000nearest}. A donor value for the current missing value is generated by sampling or aggregating the values from the $k$ nearest neighbors. In the current analysis, we identified 10 nearest neighbors using Gower's distance \cite{gower}. When imputing a single dataset using KNN, we aggregated values from nearest neighbors using the median for numeric variables and the mode for categorical variables. When imputing multiple datasets using KNN, we sampled one value at random from nearest neighbors for each imputed set.

@@

The Brier score is dependent on the rate of observed events, which can make it a difficult metric to interpret. It is often more informative to scale the Brier score based on the Brier score of a naive model. More specifically, for a given risk prediction model, the scaled Brier score is computed as $$\textrm{Scaled } \bstat \textrm{ of model } = 1 - \frac{\bstat \textrm{ of model}}{\bstat \textrm{ of naive model}}.$$ As the Brier score for risk prediction is analogous to mean-squared error for prediction of a continuous outcome, the scaled Brier score is analogous to the $R^2$ statistic. Similar to the $R^2$ statistic, a scaled $\bstat$ of 1.00 and 0.00 indicate a perfect and worthless model, respectively. In our analyses, a Kaplan-Meier estimate based on the training data (\ie a risk prediction model that did not use any predictor variables) provided the naive prediction.

@@

To assess the prognostic value of each missing data strategy, we internally validated a total of FILL IN modeling algorithm based on combinations of imputation strategies and modeling strategies described in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (\eg imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) \cite{kuhn2013applied}. We conducted internal validation using 200 replicates of Monte-Carlo cross validation (MCCV), a resampling technique for internal validation (Figure FILL IN).

\paragraph{Steps taken in each MCCV replicate} In each replicate, 50\% of the available data were used for model training and testing. Imputation was conducted in the training and testing sets, separately, for each imputation strategy. Some imputation strategies (\eg KNN and random forests) used only the training data to generate imputed values, whereas others (\eg Bayesian regression, PMM, and hot deck)

After imputation, each modeling algorithm (\ie stepwise PH models, GBDTs, and the cIF) was applied, separately, to develop a RPE for death or transplant that incorporated up to 50 predictor variables from the entire collection of covariates available.

@@

Participant characteristics were tabulated for the overall population and stratified by event status. Continuous and categorical variables were summarized as mean (standard deviation) and percent, respectively. Imputation accuracy was aggregated for all numeric and nominal variables to create two overall scores for each imputed dataset, and the distribution of these scores was tabulated for each imputation strategy.

@@

implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}. \texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable. The procedure may repeat until estimated generalization error of the random forest collection stabilizes.

@@

trunc_events <- im %>% 
  summarize(
    n_dead = sum(pt_outcome_dead == 1 & months_post_implant < times),
    n_txpl = sum(pt_outcome_txpl == 1 & months_post_implant < times),
    n_cens = sum(pt_outcome_cess == 0 & 
                 pt_outcome_dead == 0 &
                 pt_outcome_txpl ==0 & 
                 months_post_implant < times),
    p_dead = 100 * mean(pt_outcome_dead == 1 & months_post_implant < times),
    p_txpl = 100 * mean(pt_outcome_txpl == 1 & months_post_implant < times),
    p_cens = 100 * mean(pt_outcome_cess == 0 & 
                 pt_outcome_dead == 0 &
                 pt_outcome_txpl ==0 & 
                 months_post_implant < times)

  ) %>% 
  transmute(
    dead = table_glue("{n_dead} ({p_dead}%)"),
    txpl = table_glue("{n_txpl} ({p_txpl}%)"),
    cens = table_glue("{n_cens} ({p_cens}%)")
  ) %>% 
  as.list()

@@

@@

\paragraph{Imputation accuracy}

>>>>

@@

drake::loadd(im, 
             times, 
             rspec,
             md_method_labels,
             md_type_labels,
             model_labels,
             outcome_labels,
             additional_missing_labels,
             tbl_descriptives,
             tbl_impute_accuracy,
             tbl_md_strat,
             fig_md_strat_infer,
             gt_tbls)

tbl_characteristics <- tbl_descriptives$characteristics
tbl_missingness <- tbl_descriptives$missingness

@@

The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nation wide level. Currently, INTERMACS comprises over 20,000 patients who have received a MCS device. As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step for developing risk prediction models is handling missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial. While some missing values are related to data entry error, others may relate to instability of a patient’s circulatory status at the time of implant.  For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed. Patients without invasive catheterizations may be too ill (unstable) to tolerate the procedure. Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.

@@

In this section, we describe the INTERMACS Registry in Section \ref{subsec:intermacs} and provide a broad overview of general objectives for imputation of missing data in the context of statistical inference and supervised statistical learning in Section \ref{subsec:inference_and_learning}. We then describe thirteen strategies to handle missing values (Section \ref{subsec:imputation}) and three strategies to develop risk prediction models after missing data have been imputed (Section \ref{subsec:modeling}). Next, we describe how risk prediction models and by extension the imputation strategy applied before training the model were evaluated in Section \ref{subsec:evaluation}. We outline an internal validation procedure that was applied to compare combinations of imputation and modeling strategies using hierarchical Bayesian models in Section \ref{subsec:internal}. Last, details on computation and software packages used for the current analysis are described in Section \ref{subsec:computing}

@@

The INTERMACS data is publicly available on biolincc at https://biolincc.nhlbi.nih.gov/studies/intermacs/. INTERMACS is a North American observational registry for patients receiving MCS devices that began as a partnership between the National Heart, Lung, and Blood Institute, US Food and Drug Administration, the Centers for Medicaid and Medicare Services, industry, and individual hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS became an official Society of Thoracic Surgeons database.

@@

\paragraph{Using outcomes during imputation} The outcome variable should be used to impute predictor values for statistical inference, but using outcome variables for this purpose in supervised learning can have unintended consequences in model implementation. For instance, suppose we seek to predict an outcome $Y$ for a patient in the clinical setting. If the patient is missing information for a predictor $X$ and our model's strategy to impute missing values of $X$ leverages the observed value of $Y$, how should we impute $X$? If we do not already know $Y$, then we cannot impute $X$ and hence cannot generate a valid prediction. On the other hand, if we already know $Y$, then we do not need to predict it. Due to these practical considerations and because INTERMACS is often leveraged to create prediction models for clinical practice \cite{thomas2014pre}, we did not leverage outcome variables to impute missing values of predictors in the current analysis.

@@

\paragraph{K-Nearest-Neighbors (KNN)} KNN imputation identifies $k$ `similar' observations (\ie neighbors) for each observation with a missing value in a given variable \cite{chen2000nearest}. A donor value for the current missing value is generated by sampling or aggregating the values from the $k$ nearest neighbors. In the current analysis, we identified 10 nearest neighbors using Gower's distance \cite{gower}. When imputing a single dataset using KNN, we aggregated values from 10 nearest neighbors using the median for numeric variables and the mode for categorical variables. When imputing multiple datasets using KNN, we aggregated values from 2, 6, 11, 16, and 20 neighbors, separately, to create 5 imputed datasets.

@@

The Brier score is dependent on the rate of observed events, which can make it a difficult metric to interpret. It is often more informative to scale the Brier score based on the Brier score of a naive model. More specifically, for a given risk prediction model, the scaled Brier score is computed as $$\textrm{Scaled } \bstat \textrm{ of model } = 1 - \frac{\bstat \textrm{ of model}}{\bstat \textrm{ of naive model}}.$$ As the Brier score for risk prediction is analogous to mean-squared error for prediction of a continuous outcome, the scaled Brier score is analogous to the $R^2$ statistic. Similar to the $R^2$ statistic, a scaled $\bstat$ of 1.00 and 0.00 indicate a perfect and worthless model, respectively. In our analyses, a Kaplan-Meier estimate based on the training data (\ie a risk prediction model that did not use any predictor variables) provided the naive prediction. In the current analysis, we multiply scaled Brier score values by 100 for ease of interpretation.

\paragraph{Discrimination and calibration}

The discrimination of a risk prediction model measures the probability that the model will successfully identify which of two observations is at higher risk for the event of interest. We estimated discrimination using a time-dependent concordance (C-) index that accounted for covariate-dependent censoring. A C-index of 0.50 and 1.00 correspond to worthless and perfect discrimination, respectively. Similar to the scaled Brier score, we multiply C-index values by 100 for ease of interpretation. Calibration slope plots measure a risk prediction model's absolute accuracy. We estimated calibration error by averaging the squared distance between expected and observed event rates according to a calibration plot. Full description of these evaluation metrics are available \cite{gerds2014calibration, gerds2013estimating}.

@@

To assess the prognostic value of each missing data strategy, we internally validated a total of 23 modeling algorithm based on combinations of imputation strategies and modeling strategies described in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For convenience, we use the term `modeling algorithm' to denote the combination of a missing data strategy and a modeling strategy (\eg imputation using mean/mode values followed by fitting the PH model with stepwise variable selection) \cite{kuhn2013applied}. We conducted internal validation using 200 replicates of Monte-Carlo cross validation, a resampling technique for internal validation.

\paragraph{Steps taken in each replicate} In each replicate of Monte-Carlo cross validation, 50\% of the available data were used for model training and testing. All predictor variables with < 50\% missing values were considered for imputation and subsequent model development. Among these variables, artificial missingness (0\%, 15\%, or 30\% additional missing values) was induced based on patient age, with younger and older patients more likely to have missing data compared to patients who were between 40 and 65 years of age. Prior to imputation, 50 predictor variables were selected using a boosting model that quantified variable importance as the fractional contribution of each predictor to the model based on the total gain attributed to using the predictor while growing decision trees. Imputation was conducted in the training and testing sets, separately, for each imputation strategy. Although some imputation strategies (\eg KNN and random forests) can impute data in the testing set using models fitted to the training set, others (\eg Bayesian regression, PMM, and hot deck) cannot. Therefore, to ensure fair comparisons in our experiment, each imputation procedure imputed data in the training set using models fitted to the training set and then imputed data in the testing set using models fitted to the testing set. After imputation, Cox PH and boosting models were applied to each imputed dataset, separately. Last, model predictions for death and transplant were computed at 6 months following MCS surgery.

\paragraph{Bayesian analysis of model performance} To determine whether any of the imputation methods described above had improved upon imputation to the mean, we applied Bayesian hierarchical models to analyze differences in scaled Brier score \cite{benavoli2017time}. This strategy provides a flexible framework to conduct hypothesis testing and also accounts for correlation occurring within each replicate of Monte-Carlo cross validation. Specifically, within each replicate of Monte-Carlo cross validation, the performance of different modeling algorithms are correlated because they are trained and tested using the same data.

@@

Participant characteristics were tabulated for the overall population and stratified by event status. Continuous and categorical variables were summarized as mean (standard deviation) and percent, respectively. The number and proportion of missing values were also tabulated for the overall population and stratified by event status. Missing data patterns were visualized for the overall population using an UpSet plot \cite{lex2014upset}. All of the proceeding analyses were conducted using resampling results from Monte-Carlo cross validation. Imputation accuracy was aggregated for all numeric and nominal variables to create two overall scores for each imputed dataset. For imputation methods that created multiple datasets, scores were aggregated over each dataset to provide a single summary score. The distribution (\ie 25^th^, 50^th^, and 75^th^ percentile) of the scaled Brier score was estimated for each modeling algorithm.

We split our results from Monte-carlo cross validation into four datasets based on outcome (mortality or transplant) and modeling procedure (Cox PH or boosting). For each dataset, we fit one hierarchical Bayesian model where the dependent variable was scaled Brier score and independent variables included the imputation strategy applied and the amount of artificially missing data (0\%, 15\%, or 30\%) induced before imputation. With each model, we estimated the difference in scaled Brier score of downstream models when random forests, Bayesian regression, PMM, hot deck, MIA, or KNN imputation were applied to impute missing values instead of imputation to the mean.

@@

Our source code is available on GitHub (see [https://github.com/bcjaeger/INTERMACS-missing-data](https://github.com/bcjaeger/INTERMACS-missing-data)).

<!-- We  implement \texttt{miceRanger}, an R package similar to \texttt{missForest} that implements MICE with random forests by implementing PMM \cite{missForest}. \texttt{miceRanger} follows an iterative procedure where one random forest is fit to each variable in a dataset, separately, and then used to impute missing values in that variable. The procedure may repeat until estimated generalization error of the random forest collection stabilizes.  -->

@@

trunc_events <- im %>% 
  summarize(
    n_dead = sum(pt_outcome_dead == 1 & months_post_implant < times),
    n_txpl = sum(pt_outcome_txpl == 1 & months_post_implant < times),
    n_cens = sum(pt_outcome_cess == 0 & 
                 pt_outcome_dead == 0 &
                 pt_outcome_txpl ==0 & 
                 months_post_implant < times),
    p_dead = 100 * mean(pt_outcome_dead == 1 & months_post_implant < times),
    p_txpl = 100 * mean(pt_outcome_txpl == 1 & months_post_implant < times),
    p_cens = 100 * mean(pt_outcome_cess == 0 & 
                 pt_outcome_dead == 0 &
                 pt_outcome_txpl ==0 & 
                 months_post_implant < times)
  ) %>% 
  transmute(
    dead = table_glue("{n_dead} ({p_dead}%)"),
    txpl = table_glue("{n_txpl} ({p_txpl}%)"),
    cens = table_glue("{n_cens} ({p_cens}%)")
  ) %>% 
  as.list()

@@

inline_missingness <- as_inline(tbl_missingness,
                                tbl_variables = c('status', 'variable'),
                                tbl_value = 'tbl_value')

@@

\paragraph{Missing data} Many predictor variables exhibited similar proportions of missing values in different outcome groups (Table \ref{tbl_missingness}). However, the number (percent) of missing values for surgery time was an exception, with `r inline_missingness$Overall$im_surgery_time` in the overall population, `r inline_missingness$Dead$im_surgery_time` among patients who died, and `r inline_missingness$Censored$im_surgery_time` among patients who were censored. Additionally, missing values for surgery time were frequently accompanied by missing values for CV pressure (Figure \ref{fig:upset}).

```{r inline_impute_accuracy}

si_higher_accuracy <- tbl_impute_accuracy %>% 
  drop_na() %>% 
  summarize(
    n_nominal = sum(nominal_est_si > nominal_est_mi),
    n_numeric = sum(numeric_est_si > numeric_est_mi),
    n_total = n()
  )

top_accuracy <- tbl_impute_accuracy %>% 
  select(-c(nominal_si, nominal_mi, numeric_si, numeric_mi)) %>% 
  pivot_longer(cols = -c(md_method, additional_missing_pct),
               names_to = c('score', 'estimand', 'md_type'),
               names_sep = '_',
               values_to = 'value') %>% 
  filter(estimand == 'est') %>% 
  group_by(md_method, additional_missing_pct, score, md_type) %>% 
  summarize(value = mean(value), .groups = 'drop') %>% 
  arrange(desc(value)) %>% 
  group_by(score, additional_missing_pct) %>% 
  slice(1) %>% 
  left_join(tbl_impute_accuracy) %>% 
  ungroup() %>% 
  split(list(.$score, .$additional_missing_pct)) %>% 
  map(select, md_method, nominal_si, numeric_si)

break_ci <- function(x, pre_label){

  data <- strsplit(x, split = ' \\(') %>% 
    unlist() %>% 
    str_remove('\\)') %>% 
    set_names(c('estimate', 'interval')) %>% 
    as.list()

  glue("{pre_label}{data$estimate}, 95% CI {data$interval}")

\paragraph{Imputation accuracy} Single imputation strategies were more accurate than their counterparts using multiple imputation in `r si_higher_accuracy$n_nominal` out of `r si_higher_accuracy$n_total` comparisons of nominal scores and `r si_higher_accuracy$n_numeric` out of `r si_higher_accuracy$n_total` comparisons of numeric scores (Table \ref{tbl_impute_accuracy}). When an additional 15% of data were missing, single imputation KNN obtained the highest nominal accuracy (`r break_ci(top_accuracy$nominal.15$nominal_si, pre_label = 'score: ')`) and single imputation with random forests obtained the highest numeric accuracy (`r break_ci(top_accuracy$numeric.15$numeric_si, pre_label = 'score: ')`). When an additional 30% of data were missing, imputation to the mean obtained the highest numeric and nominal scores.

```{r inline_md_strat}

inline_md_strat <- tbl_md_strat %>%
  bind_rows(.id = 'outcome') %>% 
  mutate(tbv = table_glue("{est} ({lwr}, {upr})", rspec = rspec)) %>% 
  split(f = list(.$outcome,
                 .$model,
                 .$md_strat,
                 .$additional_missing_pct)) %>% 
  map(pull, tbv)

## Scaled Brier score

\paragraph{Mortality risk prediction}

When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$dead.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.meanmode_si.0` for boosting models (Table \ref{tbl_md_strat_dead_ipa}; top panel). Multiple imputation strategies universally obtained higher scaled Brier scores versus their single imputation counterparts. Multiple imputation using random forests provided the highest scaled Brier score compared to other strategies, leading to a median (25^th^, 75^th^ percentile) increase in the scaled Brier score of `r inline_md_strat$dead.ipa.cph.ranger_mi.0` for Cox PH and `r inline_md_strat$dead.ipa.xgb.ranger_mi.0` for boosting models. These performance increments improved when an aditional 15\% and 30\% of data were amputed (Table \ref{tbl_md_strat_dead_ipa}; middle and bottom panel).

\paragraph{Transplant risk prediction}

When no additional data were amputed and imputation to the mean was applied before fitting risk prediction models, the median (25^th^, 75^th^ percentile) scaled Brier score was `r inline_md_strat$txpl.ipa.cph.meanmode_si.0` for Cox PH and `r inline_md_strat$txpl.ipa.xgb.meanmode_si.0` for boosting models (Table \ref{tbl_md_strat_txpl_ipa}; top panel). For Cox PH models, imputation to the mean provided the lowest scaled Brier score, and random forest imputation led to a median (25^th^, 75^th^ percentile) increase of `r inline_md_strat$txpl.ipa.cph.ranger_mi.0` versus imputation to the mean. For boosting models, surprisingly, imputation to the mean provided a higher scaled Brier score than all imputation methods except for multiple imputation using Bayesian regression. When an additional 15\% and 30\% of data were amputed, the performance increments corresponding to the use of multiple imputation increased for both Cox PH and boosting models (Table \ref{tbl_md_strat_txpl_ipa}; middle and bottom panel).

## Discrimination and calibration

\paragraph{Mortality risk prediction}

Regardless of how much additional data were amputed, boosting models obtained higher median C-indices than PH models for prediction of 6-month mortality risk (Table \ref{tbl_md_strat_dead_auc}). In addition, all models that used multiple imputation consistently obtained higher median C-indices compared to their counterparts using single imputation. When 0 and 15\% additional data were amputed, median calibration error was lower for PH models compared to boosting models, but boosting models obtained lower median calibration error when 30\% additional data were amputed (Table \ref{tbl_md_strat_dead_cal_error}). Almost all imputation strategies provided lower median calibration error compared with imputation to the mean.

\paragraph{Transplant risk prediction}

For all amounts of additional data amputed, PH models obtained higher median C-indices than boosting models for prediction of 6-month transplant risk (Table \ref{tbl_md_strat_txpl_auc}). Similar to mortality risk prediction, multiple imputation strategies generally provided higher C-indices than their counterparts using single imputation strategy. For boosting models, MIA provided higher C-indices compared to all other single imputation strategies and had similar or superior performance compared to several multiple imputation strategies. Differences in calibration error were minor when no additional missing data were amputed (Table \ref{tbl_md_strat_txpl_cal_error}). When an additional 30\% of data were amputed, boosting models using MIA obtained lower calibration error than any other strategy.

\paragraph{Bayesian analysis of model performance}

```{r}

fig_md_strat_inline <- fig_md_strat_infer %>% 
  map(~ map(.x$probs, as_inline, 'm2', 'prob_gt_meanmode'))

inline_ipa_rf <- fig_md_strat_inline$ipa$`Multiple imputation`$ranger
inline_auc_rf <- fig_md_strat_inline$auc$`Multiple imputation`$ranger
